---
title: "Project 1"
author: "Fernando Villarroel"
---
   
This assignment is answering the questions detailed in the questionary. it is completed in a single R markdown document, and later processed by knitr and transformed into an HTML file.
   
####Loading and preprocessing the data
The data is loaded in a data.frame with the name of ***activity***, the Class of every column is assigned at this moment:

```{r LoadData}
activity<-read.csv("activity.csv",colClasses=c("numeric","POSIXct","numeric"))
```
####What is mean total number of steps taken per day?
  
Firt is calculated the total steps taken per day and it is saved in a data.frame with the  name ***DailyActSumDF***.  
  
The **mean** and **median** total number of steps taken per day are calculated and stored in the variables ***DailyActMean*** and ***DailyActMedian*** respectivily.  

```{r DailyActivity}
DailyActSum<-tapply(activity$steps,activity$date,sum,na.rm=F)
DailyNames<-names(DailyActSum)
DailyActSum<-as.numeric(DailyActSum)
DailyActSumDF<-data.frame(Date=DailyNames,TotalSteps=DailyActSum)
DailyActMean<-round(mean(DailyActSum,na.rm=T),0)
DailyActMedian<-round(median(DailyActSum,na.rm=T),0)
```
  
The Plot is builded using **ggplot2***.  

```{r HistDailyActivity,fig.height=4}

library(ggplot2)

gh<-ggplot(DailyActSumDF,aes(x=TotalSteps))+
  geom_histogram(binwidth=2000,color="blue",fill="green")+
  ggtitle("Histogram of the total steps per day \n ") +         
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  labs(x = "Total Steps per day", y = "Frecuency")
  
print(gh)
png("./figures/1-DailyActSum.png")
print(gh)
dev.off()

```
  
The the mean total number of steps taken per day is: **`r DailyActMean`**.  
The the median total number of steps taken per day is **`r DailyActMedian`**.  
  
####What is the average daily activity pattern?
  
Firt is calculated the average steps taken per interval across all the days and it is saved in a data.frame with the  name ***IntervalActMeanDF***.  
  
The interval in which is taken the maximun numbers of steps is calculates and stored in the variable **MaxInterval**.  

```{r IntervalyActivity}
IntervalActMean<-tapply(activity$steps,activity$interval,mean,na.rm=T)
IntervalNames<-names(IntervalActMean)
IntervalActMean<-as.numeric(IntervalActMean)
IntervalActMeanDF<-data.frame(Interval=as.numeric(IntervalNames),AverageSteps=IntervalActMean)
IntervalActMax<-max(IntervalActMean,na.rm=T)
IntervalActMaxLog<-IntervalActMean==IntervalActMax
MaxInterval<-IntervalNames[IntervalActMaxLog]
```
  
The Plot is builded using **ggplot2***.  

```{r TimeSeriePlot,fig.height=4}

library(ggplot2)

gi<-ggplot(IntervalActMeanDF,aes(x=Interval,y=AverageSteps))+
  geom_point(color="blue",size=2)+
  geom_path(color="blue")+
  ggtitle("Average steps by Interval\n ") +         
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  labs(y = "Average Steps")
  
print(gi)
png("./figures/2-IntervalActMean.png")
print(gi)
dev.off()
```
  
On average across all the days in the dataset the 5-minute interval which contains the maximum number of step is the: **`r MaxInterval`**.  
  
####Imputing missing values
  
1)  To calculate the number of missing values is used the functions ***is.na*** and ***sum***, the number is stored in the variable ***isnanumber***. 
  
```{r NACalculation}
isnanumber<-sum(is.na(activity$steps))
```
  
The number of  missing values in the dataset is: **`r isnanumber`**
  
2)  The strategy for filling the missing values in the dataset is use the mean for that 5-minute interval in every missed value.

3) The new dataset that is equal to the original dataset but with the missing data filled with the strategy described in point 2 is ***ActNoNA***, to create it was used a loop.
    
```{r FillMissingValues}
ActNoNA<-activity
for(i in 1:nrow(ActNoNA)){
  isna<-is.na(ActNoNA[i,1])
  if(isna==T){
    IntLog<-IntervalNames==ActNoNA[i,3]
    ActNoNA[i,1]<-round(IntervalActMean[IntLog],0)
  }

}
```
  
4) To create the new **histogram** and calculate the new **mean** and **mediam** are used the same steps as the first task.
  
```{r DailyActNoNA}
DailyActNoNASum<-tapply(ActNoNA$steps,ActNoNA$date,sum,na.rm=F)
DailyNoNANames<-names(DailyActNoNASum)
DailyActNoNASum<-as.numeric(DailyActNoNASum)
DailyActNoNASumDF<-data.frame(Date=DailyNoNANames,TotalSteps=DailyActNoNASum)
DailyActNoNAMean<-round(mean(DailyActNoNASum,na.rm=T),0)
DifMean<-abs(DailyActNoNAMean-DailyActMean)
DailyActNoNAMedian<-round(median(DailyActNoNASum,na.rm=T),0)
DifMedian<-abs(DailyActNoNAMedian-DailyActMedian)
```
  
The Plot is builded using **ggplot2***.  

```{r HistDailyActNoNA,fig.height=4}

library(ggplot2)

ghnona<-ggplot(DailyActNoNASumDF,aes(x=TotalSteps))+
  geom_histogram(binwidth=2000,color="blue",fill="green")+
  ggtitle("Histogram of the total steps per day \n ") +         
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  labs(x = "Total Steps per day", y = "Frecuency")
  
print(ghnona)
png("./figures/3-DailyActNoNASum.png")
print(gh)
dev.off()
```
  
The new mean total number of steps taken per day is: **`r DailyActNoNAMean`**, the difference with the previous value is **`r DifMean`** steps.  
   
The new median total number of steps taken per day is **`r DailyActNoNAMedian`**, the difference with the previous value is **`r DifMedian`** steps.  
   
The main differences with the data sets with missing values and the data set without missing values is that the frequency of the average values increase around 10 points.
   
####Are there differences in activity patterns between weekdays and weekends?
   
1)  A new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day has been create, the column has the name of ***TypeDay***.
   
```{r WeekDay}
Sys.setlocale("LC_TIME", "C")
ActNoNA$weekday<-as.factor(weekdays(ActNoNA$date))
TypeDay<-character()
for(i in 1:nrow(ActNoNA)){
  if(ActNoNA[i,4]=="Monday"|ActNoNA[i,4]=="Tuesday"|ActNoNA[i,4]=="Wednesday"|ActNoNA[i,4]=="Thursday"|ActNoNA[i,4]=="Friday"){
    TypeDay[i]<-"Weekday"
  } else {TypeDay[i]<-"Weekend"}
}
ActNoNA$TypeDay<-TypeDay
```

2.  First the data is prepare in a data set in which the first column has the 5-minute interval -***Interval***- the second column has the average for that interval depending of the type of day -***Mean***- and the third column has the type of day, Weekday or Weekend -***TypeDay***-.

```{r IntervalyActivityperTypeDay}
IntercalActNoNASTD<-split(ActNoNA,ActNoNA$TypeDay)

IntervalActNoNAMeanWD<-tapply(IntercalActNoNASTD[[1]]$steps,IntercalActNoNASTD[[1]]$interval,mean,na.rm=T)
IntervalNoNANamesWD<-names(IntervalActNoNAMeanWD)
IntervalActNoNAMeanWD<-as.numeric(IntervalActNoNAMeanWD)
IntervalActNoNAMeanDFWD<-data.frame(Interval=as.numeric(IntervalNoNANamesWD),AverageSteps=IntervalActNoNAMeanWD,TypeDay="Weekday")

IntervalActNoNAMeanWE<-tapply(IntercalActNoNASTD[[2]]$steps,IntercalActNoNASTD[[2]]$interval,mean,na.rm=T)
IntervalNoNANamesWE<-names(IntervalActNoNAMeanWE)
IntervalActNoNAMeanWE<-as.numeric(IntervalActNoNAMeanWE)
IntervalActNoNAMeanDFWE<-data.frame(Interval=as.numeric(IntervalNoNANamesWE),AverageSteps=IntervalActNoNAMeanWE,TypeDay="Weekend")

IntervalActNoNAMeanDF<-rbind(IntervalActNoNAMeanDFWD,IntervalActNoNAMeanDFWE)

```
  
The Plot is builded using **ggplot2***.  

```{r TimeSeriePlotperTypeDay,fig.height=4}

library(ggplot2)

gtd<-ggplot(IntervalActNoNAMeanDF,aes(x=Interval,y=AverageSteps))+
  geom_point(color="blue",size=1)+
  geom_path(color="blue")+
  ggtitle("Average steps by Interval and type of day\n ") +         
  facet_grid(TypeDay~.)+
  theme(plot.title = element_text(lineheight=.8, face="bold"))+
  labs(y = "Average Steps")
  
print(gtd)
png("./figures/4-IntervalActNoNAMean.png")
print(gh)
dev.off()

```